---
title: batch convert docx documents with images to markdown using R + pandoc
author: Thomas Goossens
date: '2020-07-13'
categories:
  - r
  - docker
  - linux
  - shiny
tags:
  - how-to
---

Do you or your organization have a batch a MS Word `.docx` files that need to be converted to markdown ? 

In this post I'll show you how you can use the power of Pandoc and RMarkdown to achieve this operation. And it will even work with `.docx` files containing tables and or images.

It's basically a small R function that will parse a specified folder and call a pandoc command for each of the `.docx` files found in this folder. Here is the source code of the function

```{r eval=FALSE}
convert_docx_to_md = function(folder){
  docx_files_path = as.list(list.files(path = folder, pattern = ".docx"))
  docx_files_names = as.list(list.files(path = folder, full.names = FALSE, pattern = ".docx"))

  convert = function(docx_file_name){
    
    docx_file_path = paste0(
      folder, 
      "/",
      docx_file_name
    )
    
    file_name = gsub(pattern = "\\.docx$", "", docx_file_name)
    
    dir.create(
      path = paste0(
        folder,
        "/",
        file_name
      ))
    
    file.copy(
      from = paste0(
        folder,
        "/",
        file_name, ".docx"
      ),
      to = paste0(
        folder,
        "/",
        file_name,
        "/",
        file_name,
        ".docx"
      )
    )
    
    command_pandoc = paste0(
      "pandoc -f docx -t markdown_mmd ",
      folder,"/",
      file_name, "/", file_name,
      ".docx ",
      paste0("--output=", folder, "/", file_name, "/", file_name, ".md"),
      paste0(" --extract-media=", folder, "/", file_name, "/")
    )
    system(command = command_pandoc)
    file.remove(paste0(
      folder,
      "/",
      file_name,
      "/",
      file_name,
      ".docx"
    ))
  }

  lapply(docx_files_names, function(x)convert(x))
}
```

To use it, copy and paste it in a `.R` file that you name e.g. `docx_to_md.R` and save in your working directory. Then source the file : 

```{r eval=FALSE}
source("docx_to_md.R")
```

Now, you can call the function by passing it the full path of the folder containing the `.docx` documents. Let's pretend you want to convert a batch of `.docx` files stored into the `files_to_convert` folder present on your Desktop. Here is the 2 lines of code that you need to execute to convert all the files :

```{r eval=FALSE}
folder = "C:/Desktop/files_to_convert"
convert_docx_to_md(folder = folder)
```

As markdown files are plain text files, the images are not stored in the file itself. The file simply contains a link that points to the image file to be displayed. To avoid names collapses, Each of the images that must be rendered in a markdown file must therefore be stored in a folder

What does the function do ? 

The function take the docx files one by one and for each file, it will proceed as follows : 

1. copy the `.docx` file in a new subfolder that takes the exact same name as the `.docx` file. 
* in this subfolder, convert the `.docx` to a markdown file that takes the exact same name as the copied `.docx` file. If this `.docx` file document contains images, the function will create a subfolder named `media` into which all the images of the `.docx` file will be stored. This is required to avoid 